The Courtroom Challenge—A Conjecture on the Unavoidable Cost of LLM Verification

### 2025-06-29 paper/main.tex
Add files via upload

To address these shortcomings we pose the \textit{Courtroom Challenge}: a
hybrid state machine in which every LLM claim is tested against a persistent
precedent store and—when automatic checks falter—escalated under a
tunable \textit{validation\_preference} dial (formal spec in
Section~\ref{sec:framework}).  The design rewires known parts
(generator, retrieval, human oversight) but prices each byte, cycle,
and minute of validation.  Retrieval-augmented generation \citep{lewis2020rag,
fan2024ragsurvey}—and later iterative variants—motivate the precedent
store; human-alignment work such as Constitutional AI
\citep{bai2022constitutional} motivates the escalation path.
Unlike debate frameworks \citep{chen2024reconcile,madaan2023selfrefine} or formal
provers \citep{wang2025malot}, we emphasise continuous cross-session
verification: knowledge is never final, only current.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\linewidth]{courtroom}%
      % no .pdf extension needed if LaTeX finds courtroom.pdf
  \caption{One step of the Courtroom loop.%
           \ Solid arrows: primary query flow.%
           \ Dashed arrows: helper calls and cost logging.%
           \ The cost\,bus aggregates all resource writes into a single ledger.
           \ On \textsc{DEFER} the round is skipped (no cost, no memory write).
           \ If C cannot supply an authoritative value—because the fact is genuinely unknown—the 
           \ query transitions to a terminal UNRESOLVED state; no memory write occurs.}
  \label{fig:courtroom}
\end{figure}

% ---------- quick-reference box ------------------------------------
\begin{tcolorbox}[colback=black!3,colframe=black!40,
  boxrule=0.4pt, arc=2pt, left=2pt, right=2pt,
  top=4pt, bottom=4pt, title=Notation snapshot]
\small
\begin{itemize}[leftmargin=*]
\item $G$ – \textbf{Generator}: LLM (+ optional helper calls)
that proposes a hypothesis $h_t$ for query $q_t$.
\item $V_{\theta}$ – \textbf{Validator} with
\emph{validation–preference dial} $\theta\!\in[0,1]$  
($0 =\,$speed‐first, $1 =\,$accuracy‐first).  
The dial is fixed during any evaluation run; see
Sec.~\ref{sec:framework}.
\item $\kappa_i c_{i,t}$ – \textbf{Cost-ledger term}: each resource $i$
(bytes written, CPU‐seconds, human minutes …) is logged as a
quantity $c_{i,t}$ weighted by its unit cost $\kappa_i$.
$\sum_i\kappa_i c_{i,t}$ is the \emph{friction} paid at step $t$.
\item $\tau$ – \textbf{Retention threshold} that controls whether a
new descriptor vector is written to the precedent store:  
\vspace{-2pt}
\[
\text{write\_new}=\begin{cases}
[\cos\theta < \tau], &\text{(cosine form)}\\[4pt]
[\lVert x-y\rVert_2 > \tau], &\text{(distance form)}
\end{cases}
\]
Designers may substitute any monotone similarity metric.
\end{itemize}
\vspace{2pt}\par\footnotesize (Formal definitions → Section~\ref{sec:framework})\par
\end{tcolorbox}
% --------------------------------------------------------------------

@@

\paragraph{3. External judge $C$}
A human analyst consults the 2018 10-K and confirms that
\texttt{Debt Securities} is a \emph{new} line-item, not yet present in
the taxonomy.  The analyst extracts the FY-2018 amount  
\[
a_t = \text{``Bucket: \texttt{"Debt Securities"}; FY-2018 amount: \$14.406 B.''}
\]

@@

\section{Are LLMs a Dead End? Converging Error}
Critics such as LeCun \citep{LeCun2023SFI}\,%
(see also Appendix~\ref{fig:lecun-slide}) argue that
purely autoregressive LLMs will always accumulate error:
if the per-token slip rate is $e$, the probability an
$n$-token answer is entirely correct decays as
$P_{\mathrm{correct}} = (1-e)^{n}$.
LeCun also notes the inability of LLMs to represent continuous high-dimensional spaces that characterize the real world, potentially limiting their ability to deal with complex, nonlinear data.

@@

\section{Open Questions}
\label{sec:open-questions}
\begin{enumerate}
    \item Is abstraction a distinct step in hybrid reasoning? When no exact precedent exists, does the model create a new one by compressing multiple associations? Should pattern recognition and abstraction be formal components in systems like this one?
    \item Can LLMs + real-world observation produce entirely new theories? If so, how do we detect genuine novelty, and how do we measure validity?
    \item Can the Courtroom Model generalize to other real-world domains and AI integrations? Healthcare, policy, engineering — any area that needs scale, creativity, and robust checks. Could this framework point toward a general pattern for hybrid AI?
    \item How can we achieve Lean-level formal rigor in open-text domains without relying solely on large LLM ensembles, whose granularity still falls short in many fields?
    \item \label{open-question-schema} How should the precedent store be schematised and evolved? Even in a well-defined vertical like finance, designing the descriptor schema (e.g.\ `(ticker, canonical\_metric, period, unit, GAAP\_flag, source\_tag, …)`) remains an art: too coarse results in semantic collisions; too fine results in unlimited vector growth. Can the system learn an optimal schema online—merging rarely-queried fields, splitting ambiguous ones, and doing so without breaking L$_2$-distance semantics? What signals (query regret, type-confusion rate, index rebuild cost) best drive such schema evolution?
    \item How do we maintain the tension between speed and scrutiny? Associative models want to generate rapidly; symbolic structures demand thorough verification. Speed wins until it fails. Scrutiny survives because it must.
    \item What does success look like in a system that's always evolving? As new precedents form, do we prune old ones? How do we adapt long-standing truths? The answer may lie in how far our energy limitations let us revisit and revise what we already hold to be true.
    \item Which self-play or adversarial-debate schemes could fill the \(V_\theta\) slot while keeping \(\kappa_{\text{compute}}\) at or below the current human-review baseline?
\end{enumerate}
\begin{tcolorbox}[colback=gray!5,colframe=gray!40,title=Why a concept note?]
  The field currently optimises ever-larger generators without a matching theory of
  validation.  We publish this abstraction \emph{before} a prototype for two reasons:
  (i)~to surface hidden degrees of freedom that can silently invalidate benchmarks,
  and (ii)~to recruit empirical collaborators—open-sourcing code is planned for v2.
\end{tcolorbox}

@@

  \item \textbf{Memory update rule.}  
  \(U : (M, q, h, a) \to M'\) adds or merges precedent after a claim is accepted.

@@

Let $c_{i,t}$ be the measured consumption of resource $i$ at step~$t$
(bytes written, FAISS rebuild time, human minutes, \emph{etc.}).
Fix weights $\kappa_i$ once per benchmark—should the conjecture survive first contact.  The per-write cost is
\[
  \WriteCost_t \;=\; \sum_i \kappa_i\,c_{i,t}.
\]
Cumulative friction budget …  $F_T = \sum_{t=0}^{T-1}\WriteCost_t$.%
\footnote{\textsc{DEFER} incurs zero cost, so the corresponding
$\WriteCost_t$ entry is~0.}

@@

@@

\begin{lemma}[Interactive-Fano corollary]\label{lem:ErrLower}
Let $V$ be any validator run for $T$ rounds.
With $\kappa(\omega)$ and $\Err(\omega)$ defined in
\eqref{eq:kappaDef}–\eqref{eq:ErrDef}, write
\[
  \Err_{\mathrm{bar}}:=\E[\Err].
\]
Assume the transcript $X$ obeys the information budget
\(I(\theta;X)\le I\) under the \emph{uniform} prior on an $M$-point
label set ($M\ge2$).  Then
\begin{equation}\label{eq:ErrLower-new}
  \boxed{\;
     \Err_{\mathrm{bar}}
       \;\ge\;
       \frac{\delta\,T}{4}\;
       \kappa_{\!*}\;
       \Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr),
     \quad
     \kappa_{\!*}:=\E[\kappa\mid\Err>\tfrac12\delta\kappa T]
  }
  \, .
\end{equation}
\end{lemma}

@@

Unless these knobs are frozen, the framework is empirically \emph{unfalsifiable}.

@@

Because precedent management is the model's hallmark, a static snapshot is not enough; a
\emph{prequential} (streaming) phase measures learning velocity under a cost cap.

@@

\paragraph{Rules.}
\begin{enumerate}\itemsep=2pt
\item \textbf{Cost ledger:} each write incurs
\[
\WriteCost_t \;=\; \sum_i \kappa_i\,c_{i,t},\qquad
\text{unused terms may take }\kappa_i=0.
\]
\item \textbf{Delayed grading:} batch \(k\) is scored \emph{before} the system sees batch \(k{+}1\).
\item \textbf{Full disclosure:} publish the write log, cost ledger and final memory snapshot so any team can replay the run.
\end{enumerate}

@@

\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Axis} & \textbf{End-to-End Brain} & \textbf{Neuro-symbolic} & \textbf{Courtroom Model}\\ \hline
Core ambition & Single monolith handles everything & NN $+$ logic engine & LLM generates; external friction validates \\ \hline
Strength & No human loop & Explainability, deductive power & Halts hallucination without retraining \\ \hline
Weakness & Brittle world-model, hallucination & Heavy logic engineering & Testability collapses if knobs are free \\ \hline
Ideal use case & Simulated-world agents & Formal maths, compliance & High-risk domains where \emph{trust} outranks latency \\ \hline
\end{tabular}
}
\end{center}

@@

\item \emph{“Isn't this just a dressed-up retrieval-augmented loop?”} Retrieval-augmented generation (RAG) treats the store as a stateless evidence buffer flushed between prompts. The Courtroom store is jurisprudential—each write becomes precedent and must pay rent (friction) forever. That difference, though conceptually small, flips the optimisation target from “accuracy at \(k = 1\)” to “prequential regret under a cost budget.” No existing RAG benchmark measures that regime.

@@

\item \emph{“Unfalsifiable if the dial is tuned post-hoc.”}
      The evaluation protocol (Section~\ref{app:benchmark}) freezes the validation policy
      \emph{before} scoring and logs every write, so any attempt to
      retune after seeing the test set is detectable.

@@

\section*{Appendix B  \quad  Reference Slide}
\begin{figure}[h]
  \centering
  \includegraphics[width=.9\linewidth]{figs/fig_lecun_ar_llm_slide.jpeg}
  \caption{Slide from Yann LeCun's talk *“Unpopular Opinion about
           AR-LLMs”* (Santa Fe Institute, Apr 2023) \citep{LeCun2023SFI}.
           The geometric expression
           $P_{\text{correct}} = (1-e)^{n}$ motivates
           our Conjecture~\ref{conjecture1} on error accumulation in purely autoregressive
           LLMs.}
  \label{fig:lecun-slide}
\end{figure}

>>>>

To address these shortcomings we pose the \textit{Courtroom Challenge}: a
hybrid state machine in which every LLM claim is tested against a persistent
precedent store and—when automatic checks falter—escalated under a
tunable \textit{validation\_preference} dial (formal spec in
Section~\ref{sec:framework}).  The design rewires known parts
(generator, retrieval, human oversight) but prices each byte, cycle,
and minute of validation.  Retrieval-augmented generation (RAG) provides the raw materials
\citep{lewis2020rag,fan2024ragsurvey}.  
Recent advances—Self-RAG's self-verification \citep{asai2023selfrag}, 
Iter-RetGen's retrieval-generation synergy \citep{shao2023iterretgen}, 
MemoRAG's compressed long-term memory \citep{qian2024memorag}, and 
Chain-of-Verification RAG's answer-driven re-checks \citep{he2024covrag}—motivate 
our persistent precedent store but still validate only per answer.  
Human-alignment work such as Constitutional AI 
\citep{bai2022constitutional} motivates the escalation branch.  
Whereas MemoRAG retains past evidence without auditing it, our buffer makes
each accepted verdict both permanent and costly.
Unlike debate frameworks \citep{chen2024reconcile,madaan2023selfrefine} or formal
provers \citep{wang2025malot}, we emphasise continuous cross-session
verification: knowledge is never final, only current.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\linewidth]{figs/courtroom}%
      % no .pdf extension needed if LaTeX finds courtroom.pdf
  \caption{One step of the Courtroom loop.%
           \ Solid arrows: primary query flow.%
           \ Dashed arrows: helper calls and cost logging.%
           \ The cost\,bus aggregates all resource writes into a single ledger.
           \ On \textsc{DEFER} the round is skipped (no cost, no memory write).
           \ If C cannot supply an authoritative value—because the fact is genuinely unknown—the 
           \ query transitions to a terminal UNRESOLVED state; no memory write occurs.}
  \label{fig:courtroom}
\end{figure}

% ---------- quick-reference box ------------------------------------
\begin{tcolorbox}[colback=black!3,colframe=black!40,
  boxrule=0.4pt, arc=2pt, left=2pt, right=2pt,
  top=4pt, bottom=4pt, title=Notation snapshot]
\small
\begin{itemize}[leftmargin=*]
\item $G$ – \textbf{Generator}: LLM (+ optional helper calls)
that proposes a hypothesis $h_t$ for query $q_t$.
\item $V_{\theta}$ – \textbf{Validator} with
\emph{validation–preference dial} $\theta\!\in[0,1]$  
($0 =\,$speed‐first, $1 =\,$accuracy‐first).  
The dial is fixed during any evaluation run; see
Sec.~\ref{sec:framework}.
\item $\kappa_i c_{i,t}$ – \textbf{Cost-ledger term}: each resource $i$
(bytes written, CPU‐seconds, human minutes …) is logged as a
quantity $c_{i,t}$ weighted by its unit cost $\kappa_i$.
$\sum_i\kappa_i c_{i,t}$ is the friction paid at step $t$.
\item $\tau$ – \textbf{Retention threshold} that controls whether a
new descriptor vector is written to the precedent store:  
\vspace{-2pt}
\[
\text{write\_new}=\begin{cases}
[\cos\theta < \tau], &\text{(cosine form)}\\[4pt]
[\lVert x-y\rVert_2 > \tau], &\text{(distance form)}
\end{cases}
\]
Designers may substitute any monotone similarity metric.
\end{itemize}
\vspace{2pt}\par\footnotesize (Formal definitions → Section~\ref{sec:framework})\par
\end{tcolorbox}
% --------------------------------------------------------------------

@@

\paragraph{3. External judge $C$}
A human analyst consults the 2018 10-K and confirms that
\texttt{Debt Securities} is a new line-item, not yet present in
the taxonomy.  The analyst extracts the FY-2018 amount  
\[
a_t = \text{``Bucket: \texttt{"Debt Securities"}; FY-2018 amount: \$14.406 B.''}
\]

@@

\section{Are LLMs a Dead End? Converging Error}
Critics such as LeCun \citep{LeCun2023SFI}\, argue that
purely autoregressive LLMs will always accumulate error:
if the per-token slip rate is $e$, the probability an
$n$-token answer is entirely correct decays as
$P_{\mathrm{correct}} = (1-e)^{n}$.
LeCun also notes the inability of LLMs to represent continuous high-dimensional spaces that characterize the real world, potentially limiting their ability to deal with complex, nonlinear data.

@@

\section{Open Questions}
\label{sec:open-questions}
\begin{enumerate}
    \item Is abstraction a distinct step in hybrid reasoning? When no exact precedent exists, does the model create a new one by compressing multiple associations? Should pattern recognition and abstraction be formal components in systems like this one?
    \item Can LLMs + real-world observation produce entirely new theories? If so, how do we detect genuine novelty, and how do we measure validity?
    \item Can the Courtroom Model generalize to other real-world domains and AI integrations? Healthcare, policy, engineering — any area that needs scale, creativity, and robust checks. Could this framework point toward a general pattern for hybrid AI?
    \item How can we achieve Lean-level formal rigor in open-text domains without relying solely on large LLM ensembles, whose granularity still falls short in many fields?
    \item \label{open-question-schema} How should the precedent store be schematised and evolved? Even in a well-defined vertical like finance, designing the descriptor schema (e.g.\ `(ticker, canonical\_metric, period, unit, GAAP\_flag, source\_tag, …)`) remains an art: too coarse results in semantic collisions; too fine results in unlimited vector growth. Can the system learn an optimal schema online—merging rarely-queried fields, splitting ambiguous ones, and doing so without breaking L$_2$-distance semantics? What signals (query regret, type-confusion rate, index rebuild cost) best drive such schema evolution?
    \item How do we maintain the tension between speed and scrutiny? Associative models want to generate rapidly; symbolic structures demand thorough verification. Speed wins until it fails. Scrutiny survives because it must.
    \item What does success look like in a system that's always evolving? As new precedents form, do we prune old ones? How do we adapt long-standing truths? The answer may lie in how far our energy limitations let us revisit and revise what we already hold to be true.
    \item Which self-play or adversarial-debate schemes could fill the \(V_\theta\) slot while keeping \(\kappa_{\text{compute}}\) at or below the current human-review baseline?
\end{enumerate}
\begin{tcolorbox}[colback=gray!5,colframe=gray!40,title=Why a concept note?]
  The field currently optimises ever-larger generators without a matching theory of
  validation.  We publish this abstraction before a prototype for two reasons:
  (i)~to surface hidden degrees of freedom that can silently invalidate benchmarks,
  and (ii)~to recruit empirical collaborators—open-sourcing code is planned for v2.
\end{tcolorbox}

@@

  \item \textbf{Memory update rule.}
  \(U : (M, q, h, a) \to M'\) records (or merges) a precedent whenever a claim is
  accepted.  Whereas vanilla RAG treats retrieved passages as a transient
  context window \citep{lewis2020rag}—and even iterative or self-reflective
  extensions still validate only the current answer
  \citep{shao2023iterretgen,asai2023selfrag}—the Courtroom store is
  persistent: it survives across sessions and every entry can later be
  re-audited for contradiction or obsolescence.

@@

Let $c_{i,t}$ be the measured consumption of resource $i$ at step~$t$
(bytes written, FAISS rebuild time, human minutes, etc.).
Fix weights $\kappa_i$ once per benchmark—should the conjecture survive first contact.  The per-write cost is
\[
  \WriteCost_t \;=\; \sum_i \kappa_i\,c_{i,t}.
\]
Cumulative friction budget …  $F_T = \sum_{t=0}^{T-1}\WriteCost_t$.%
\footnote{\textsc{DEFER} incurs zero cost, so the corresponding
$\WriteCost_t$ entry is~0.}

@@

The definitions in \S\ref{sec:framework} merely fix notation
(\(T,\kappa,\Err,\theta,X\)); they do not restrict the validator's
internal design.  We now state Lemma~\ref{lem:ErrLower}, an
information-theoretic lower bound that applies to any interactive
validator whose transcript carries at most \(I\) bits of information
about the hidden label.  The inequality therefore functions as
motivation for the Courtroom architecture rather than a theorem
that assumes it.

@@

\begin{lemma}[Interactive-Fano corollary]\label{lem:ErrLower}
Let $V$ be any validator run for $T$ rounds.
With $\kappa(\omega)$ and $\Err(\omega)$ defined in
\eqref{eq:kappaDef}–\eqref{eq:ErrDef}, write
\[
  \Err_{\mathrm{bar}}:=\E[\Err].
\]
Assume the transcript $X$ obeys the information budget
\(I(\theta;X)\le I\) under the uniform prior on an $M$-point
label set ($M\ge2$).  Then
\begin{equation}\label{eq:ErrLower-new}
  \boxed{\;
     \Err_{\mathrm{bar}}
       \;\ge\;
       \frac{\delta\,T}{4}\;
       \kappa_{\!*}\;
       \Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr),
     \quad
     \kappa_{\!*}:=\E[\kappa\mid\Err>\tfrac12\delta\kappa T]
  }
  \, .
\end{equation}
\end{lemma}

@@

Unless these knobs are frozen, the framework is empirically unfalsifiable.

@@

Because precedent management is the model's hallmark, a static snapshot is not enough; a
prequential (streaming) phase measures learning velocity under a cost cap.

@@

\paragraph{Rules.}
\begin{enumerate}\itemsep=2pt
\item \textbf{Cost ledger:} each write incurs
\[
\WriteCost_t \;=\; \sum_i \kappa_i\,c_{i,t},\qquad
\text{unused terms may take }\kappa_i=0.
\]
\item \textbf{Delayed grading:} batch \(k\) is scored before the system sees batch \(k{+}1\).
\item \textbf{Full disclosure:} publish the write log, cost ledger and final memory snapshot so any team can replay the run.
\end{enumerate}

@@

\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Axis} & \textbf{End-to-End Brain} & \textbf{Neuro-symbolic} & \textbf{Courtroom Model}\\ \hline
Core ambition & Single monolith handles everything & NN $+$ logic engine & LLM generates; external friction validates \\ \hline
Strength & No human loop & Explainability, deductive power & Halts hallucination without retraining \\ \hline
Weakness & Brittle world-model, hallucination & Heavy logic engineering & Testability collapses if knobs are free \\ \hline
Ideal use case & Simulated-world agents & Formal maths, compliance & High-risk domains where trust outranks latency \\ \hline
\end{tabular}
}
\end{center}

@@

\item \emph{“Isn't this just a dressed-up retrieval-augmented loop?”} Retrieval-augmented generation (RAG) treats the store as a context-window scoped evidence buffer flushed between prompts. The Courtroom store is jurisprudential—each write becomes precedent and must pay rent (friction) forever. That difference, though conceptually small, flips the optimisation target from “accuracy at \(k = 1\)” to “prequential regret under a cost budget.” No existing RAG benchmark measures that regime.

@@

\item \emph{“Unfalsifiable if the dial is tuned post-hoc.”}
      The evaluation protocol (Section~\ref{app:benchmark}) freezes the validation policy
      before scoring and logs every write, so any attempt to
      retune after seeing the test set is detectable.

@@
